PassSanitaire

### 2022-03-09 ms/ms.Rmd
results done

---
title: "PassSanitaire"
author: "FD"
output: html_document
---

@@

load("../code/savevalues.RData")
```

@@

Context: The French sanitary pass led to an increase in vaccination rates in France, but spatial heterogeneities in vaccination rates are still striking [LG1]. To identify potential determinants of these heterogeneities and evaluate the the French sanitary pass' efficacy in reducing them, we used a data-driven approach on exhaustive nationwide data, gathering `r nIndicatorsAfterRemovingCorrMiss` socio-economic, political and geographic indicators.

Methods: We considered the association between being a district above the median value of the first-dose vaccination rates and being above the median value of each indicator at four time points: week `r format(as.Date(date1), "%W")`, (just before the sanitary pass announcement), week `r format(as.Date(date2), "%W")` (when the sanitary pass came fully into force), week `r format(as.Date(date3), "%W")` (end of the summer) and week `r format(as.Date(date4), "%W")` (end of data collection). We then considered the change over time of vaccination rates according to the XX most associated indicators, by deciles.

@@

Attitudes toward vaccination are also known to be influenced by social and territorial inequalities. Surveys conducted in 2020 in France showed that respondents with lower education [Schwarzinger21, Spire21], lower income levels or less trust in authorities [Spire21] were more likely to be hostile to COVID-19 vaccines. Whether differential intentions translated into effective differences in vaccination, as described in other countries [Caspi21, Murthy21], remained to be shown in France. (XX New Bajos Paper?)

@@

This study aims to obtain further insights into the association between social inequities, vaccination and sanitary pass implementation in France using nationwide data.

Methods

The French state health insurance service (Assurance Maladie) provides public datasets of exhaustive weekly first-dose vaccination data at the district scale nationally and at the suburban scale for the Paris, Lyon, and Marseille metropolitan areas. Our data included 1552 districts in mainland France (63,601,670 individuals; median district size 22,705 inhabitants; interquartile range 6,282--39,128). We associated these data with 176 socioeconomic [JA2] and 5 geographic indicators at the same geographic scales from public datasets. We allocated those indicators into categories (Activity: 12 indicators; Employment: 41 indicators; Family: 27 indicators; Geography: 5 indicators; Housing: 35 indicators; Immigration: 1 indicator; Population: 26 indicators; Revenue: 15 indicators; School: 19 indicators)

We considered the association between being a district in the lowest quartile (25th percentile) of the first-dose vaccination rate and being above the median value of each indicator. We computed odds ratios and their 95% confidence intervals at three time points: week 27 (just before the sanitary pass announcement), week 31 (when the sanitary pass came fully into force), and week 35 (end of data collection). For the two indicators achieving the highest odds ratios, we computed vaccination levels for each indicator quartile on the same dates. Analyses were done using R (v4.0.3). To enhance odds-ratio visualization, we computed Manhattan plots giving odds-ratios and their confidence intervals by socioeconomic indicators categories.
For each week, the significance threshold is calculated by performing 1000 random permutations on the vaccination rate. Then we take the maximum OR calculated over every indicator for each permutation. Finally the threshold is fixed at the 950th value of the maximum OR sorted in ascending order.

>>>>

---
title: "PassSanitaire"
author: "FD"
output:
  bookdown::html_document2:
    self_contained: no
    keep_md: true
    number_sections: no
---

@@

load("../code/savevalues.RData")
load("../code/data_indicators.RData") 
```

@@

Context: Implemented during the Summer 2021, the French "Sanitary Pass" led to an increase in vaccination rates in France, but spatial heterogeneities in vaccination rates are still striking [LG1]. It morphed into a "Vaccine Pass" in early 2022. To identify potential determinants of these heterogeneities and evaluate the French Sanitary and Vaccine Pass' efficacies in reducing them, we used a data-driven approach on exhaustive nationwide data, gathering `r nIndicatorsAfterRemovingCorrMiss` socio-economic, political and geographic indicators.

Methods: We considered the association between being a district above the median value of the first-dose vaccination rates and being above the median value of each indicator at different time points: just before the sanitary pass announcement (week `r format(as.Date(datesAnalysis[1]), "%Y-W%W")`), just before the sanitary pass came into force (week `r format(as.Date(datesAnalysis[2]), "%Y-W%W")`) and one month after (week `r format(as.Date(datesAnalysis[3]), "%Y-W%W")`), and the equivalent dates for the vaccine pass (weeks `r format(as.Date(datesAnalysis[4:6]), "%Y-W%W")`). We then considered the change over time of vaccination rates according to the XX most associated indicators, by deciles.

@@

Attitudes toward vaccination are also known to be influenced by social and territorial inequalities. Surveys conducted in 2020 in France showed that respondents with lower education [Schwarzinger21, Spire21], lower income levels or less trust in authorities [Spire21] were more likely to be hostile to COVID-19 vaccines. A survey conducted in July 2021 in France [Bajos2022] on close to 81000 participants indicated that unvaccinated respondents were more likely to have lower income and more likely to belong to racialised minorities than vaccinated ones. The survey also confirmed that mistrust in the government was fueling refusal to be vaccinated [Bajos2022].

<!-- XX MOVE INTERNATIONAL COMPARISONS IN THE DISCUSSION Whether differential intentions translated into effective differences in vaccination, as described in other countries [Caspi21, Murthy21], remained to be shown in France. (XX New Bajos Paper?) -->

@@

This study aims to obtain further insights on the socio-economic, political and geographic factor associated with vaccination rates, and to evaluate the effect of the French domestic sanitary pass, by using nation-wide, exhaustive datasets.

# Methods

## Data

<!-- Vaccination -->
The French state health insurance service (Assurance Maladie) provides public datasets of vaccination rates in France. These datasets are based on aggregated individual data on beneficiaries of the national health insurance service who received health care in the past year. These exhaustive datasets are updated weekly, and are provided at the district scale nationally (EPCI: *Établissement public de coopération intercommunale*, an administrative level gathering multiple towns or cities) and at the suburban scale for the Paris, Lyon, and Marseille metropolitan areas. For this study, we focused on Metropolitain France, because vaccination rates are much lower in oversea localities, and because determinants of vaccination rates are likely to differ in oversea localities compared to metropolitain ones. Our dataset included `r sum(nlocalities)` districts (`r nlocalities["EPCI"]` EPCI and `r nlocalities["PLM"]` districts at the suburban scale in Paris, Lyon, Marseille).

The vaccination dataset for mainland France encompasses about XX individuals (median district size XX, itq range XX). The vaccination data are available by age class: 00--19, 20--39, 40--54, 55--64, 65--74, 75 and over. Population sizes for each locality and each age class are also provided. These population sizes are updated monthly to account for individuals aging, potential residential moves, and deaths.

<!-- Other data -->

```{r, echo = FALSE}
insee.indicators.names <- names(typesDataAll)[!is.element(names(typesDataAll), c("Geography", "Elections"))]

tb <- typesDataNoCorr[!is.element(names(typesDataNoCorr), c("Geography", "Elections"))]
output.INSEE <- ""
for(i in 1:length(tb)) output.INSEE <- paste0(output.INSEE, names(tb)[i], ": n = ", tb[i], "; ")
output.INSEE <- substr(output.INSEE, 1, nchar(output.INSEE)-2)
```

We paired these vaccination data with three other datasets gathering socio-economic, political and geographic variables.

Socio-economic data are provided by the French national statistics institute (INSEE), and are available at the same administrative levels as the vaccination data. The different predictors are classified into `r length(insee.indicators.names)` categories (`r insee.indicators.names`).  
Latitude, longitude and surface data were extracted from maps datasets [ref OpenMaps]. We calculated from them four additional geographic indicators: distance to Paris, relative position along a South-East--North-West gradient, relative position along a South-West--North-East gradient, and local population density.  
Our political dataset consisted of the results of the two round of the 2017 Presidential election in France, which we aggregated to reconstitute the same administrative levels as the vaccination dataset. This political dataset contains the proportions of votes for each of the 11 candidates of the first round, 2 candidates of the second round (Macron and Le Pen), and proportion of abstention at each round.

These three datasets comprised `r nIndicatorsTot` indicators. We then removed those indicators with over `r round(thrMissingVals * 100)`% missing data, or with over `r thrCor` correlation with other indicators of the dataset, which left us with `r nIndicatorsAfterRemovingCorrMiss` indicators: `r sum(tb)` socio-economic indicators (`r output.INSEE`); `r typesDataNoCorr[is.element(names(typesDataNoCorr), c("Geography"))]` geographic indicators; `r typesDataNoCorr[is.element(names(typesDataNoCorr), c("Elections"))]` political indicators.

## Analysis

Vaccination was accessible to all adults in France after 27 May 2021. It opened to teenagers (12-17 year olds) on 15 June 2021, and to children (5-11 year olds) on 22 December 2021. Because of this differential accessibility of vaccines, and because vaccine passport rules also differed for non-adults, we excluded the 00-19 age class from our analysis, and focused on vaccination rates among 20+ year-old individuals (hereafter "adults").

For each indicator in our dataset, at each of the four chosen dates (weeks `r format(as.Date(datesAnalysis), "%Y-W%W")`), we considered the association between living in a district above the median of a that indicator and individual first-dose vaccination rates among adults. Odds ratios (OR) were computed from the output of a logistic regression. To be able to compare predictors irrespective of the direction of the effect, we considered the maximum of `OR, 1/OR` (hereafter $\overline{OR}$). Vaccination data are at the individual level, and indicator data at the district level.  
For each date, we determined a significance threshold by computing odds ratios on `r nrep` random permutations of a predictor, and identifying the value of the XX 99% percentile odd ratios ($\overline{OR}$) of these permuted data.

For representative indicators with the strongest association to vaccination data, we estimated age-adjusted vaccination rates among adults over time, for each decile of each indicator (treated as a factor). These estimations were obtained from a logistic model taking age class into account, and adult vaccination rates were computed on a standardized age distribution matching that of metropolitain France.

All analysis code is available at XXX; analyses were done in `r R.Version()$version.string`.

<!--

- **27 mai 2021 (S21)** : toutes les personnes majeures peuvent prendre rendez-vous pour se faire vacciner en centre de vaccination pour des injections avec les vaccins Pfizer-BioNTech ou Moderna à compter du 31 mai 2021.
- **15 juin 2021 (S24)** : les 12 à 17 ans inclus ans peuvent se faire vacciner avec le vaccin Pfizer-BioNTech en centre de vaccination. Le 24 juillet 2021, l'Agence européenne du médicament (EMA) a délivré une autorisation du vaccin Moderna sur lequel la HAS a rendu un avis favorable le 28 juillet.
- **12 juillet 2021 (S28)** : annonce du président le la république concernant la mise en place du pass sanitaire.
- **9 août 2021 (S32)** : mise en place de l'obligation du passe sanitaire dans les lieux concernés.
- **1er septembre 2021 (S35)** : une campagne de rappel est mise en place pour stimuler le système immunitaire des plus vulnérables : les résidents d'EHPAD et d'USLD, les personnes de 65 ans et plus vivant à domicile, les personnes souffrant de comorbidité(s), les personnes atteintes de pathologies à haut risque de forme grave, les personnes ayant été vaccinées avec le vaccin Janssen.
- **15 septembre 2021 (S37)** : l'obligation vaccinale concernerait tous les personnels des établissements de santé, établissements d'hébergement pour personnes âgées dépendantes (Ehpad) et établissements pour personnes en situation de handicap et tous les professionnels ou les bénévoles en contact avec des personnes âgées ou vulnérables, y compris à domicile. Cette mesure a fait l'objet d'un projet de loi relatif à la gestion de la crise sanitaire examiné par le Parlement depuis le 20 juillet 2021.

-->

# Results

We investigated the associations between each of the `r nIndicatorsAfterRemovingCorrMiss` indicators and first-dose vaccination rates. Two indicators were among the top five most associated one at all time points (see Figure \@ref(fig:figManhattan)): the share of local income coming from unemployment benefits (`Unemployment_Benef`) and vote for Mr Asselineau (`Asselineau`). The three other most associated indicators did not change in the later dates that we considered, and were the proportion of immigrants in the district (`Immigrant`), the district's relative position along a North-West--South-East gradient (`NO-SE`) and the proportion of overcrowded household (`Overcrowding_rate`).

GIVE NAMES OF NOT ASSOCIATED INDICATORS?

GIVE VALUES IN A TABLE?

```{r figManhattan, echo=FALSE, fig.height=9, fig.width=9, fig.cap="Manhattan plots of the Odds ratios for each of the indicator of our dataset, by date. Left column: around the Sanitary Pass implementation; right column: around the Vaccine Pass implementation. The top five odds ratios are labelled at each time point; the arrow next to the name indicates the direction of the effect. The gray rectangle corresponds to the 99% percentile of odds ratios in the permuted data; points falling in the rectangle are considered as non-significant."}
load("../data/predictors.RData")
load("../code/outLR.RData")
source("../code/2_plot-manhattan.R")
layout(matrix(1:6, ncol = 2))
# PLOT
plotManhattan(outLR, ntop = 5)
```

Our odds ratio calculations were based on a crude version of each indicator, which were dichotomized into values above or below the median of each indicator. To better visualize the effects (or lack thereof) of the sanitary and vaccine passes on vaccination rates over time, we computed age-adjusted vaccination rates over time, by decile of three of the most associated indicators, treated as factors (see Figure \@ref(fig:figOverTime)). While the Sanitary Pass led to an overall increase in vaccination rates, the Vaccine Pass did not affect the trend of change in vaccination rates. Heterogeneities in vaccination rates persisted after both types of pass, neatly ordered by decile of each indicator.

```{r figOverTime, fig.height=10, fig.width=4, fig.cap="Age-adjusted vaccination rates among adults, over time, by decile of each indicator. ", echo = FALSE}
load("../code/outTime.RData")
source("../code/2_plot-overTime.R")
layout(matrix(1:3, ncol = 1))
plotPropTime(outA, plotDates = TRUE, plotGraduations = TRUE)
```

Finally, historically under-vaccinated areas in France stand out as being less vaccinated against Covid-19, in particular the South-East region.

FIGURE MAP

# Discussion
